Efficient Parallelization of $\ell_1$-Minimization for Face Recognition \thanks{Corresponding author: . This work was partially supported by ARO MURI W911NF-06-1-0076.}

### 2010-11-01 paper/main.tex
Fixed algorithm pseudocode. also, y->b

%Motivate the paper by an overview about the scope and importance of $\ell_1$-minimization.
$\ell_1$-minimization is one of the most powerful tools to be added to the engineer's
toolbox in years.  It has strong theoretical motivation for both for use in recovering sparse
solutions to systems of linear equations, and has been shown to perform well even in the
presence of a combination of sparse and dense error.  Since solutions of linear systems
arise in practically every branch of engineering and there are very few application which
do {\em not} benefit from a robust error function, the applications are far to numerous to
itemize here.  
For many applications of $\ell_1$-minimization solving a system $b=Ax$, $A$ often has a
special structure that can be leveraged to compute portions of $A$ on the fly
as they are needed, speeding up perfomance of the minimization dramatically.
Unfortunately, there are many import applications where this does not hold true.
This paper targets applications where $\b$ and $A$ contain
image data, and thus must be loaded every time they are used, as is the 
case for many problems in computer vision.
The techniques presented in this paper target these applications, as well
as others where the dictionary is dense and unstructured. $A$ must be loaded every time
it is used, and algorithm performance becomes highly dependent on memory bandwidth, rather
that arithmetic logic.

% Narrow focus for applications to Recognition Application
The primary motivating application for this paper is face recognition for access control applications.
Recently, advances in automatic face recognition have been made by re-casting
it as a sparse representation problem.  This core of this technique consists of
re-casting recognition as a sparse representation problem. This is done by  
stacking the test image into a vector $\b$, stacking the training images for
all of the users into the columns of a matrix $A$, and solving the following
minimization problem:
\begin{equation}
\min_{\x, \e} \| \x \|_1 + \|\e\|_1 \quad \subj \quad \b = A \x + \e.
\end{equation}
This optimization promotes a solution that is sparse both in $\x$ and $\e$. 
Sparsity in $\x$ arrises from the fact that only a subset of the training images (if any)
correspond to the user in the test image.  Sparsity in $\e$ arrises from the knowledge
that small occlusions in the test image will only corrupt a small subset of the pixels \cite{Wright2009-PAMI}.
The large coefficients in $\x$ will concentrate on the correct user, and can be used as the basis of a classifier.

For this idea to work, all of the images must be aligned. An iterative alignment
algorithm for this purpose was described as part of the \cite{Wagner2009-CVPR}, which was
the first paper to expand \cite{Wright2009-PAMI} into a functioning face recognition system.
This iterative alignment routine is based on repeatedly solving the following optimization problem:
\begin{equation}
\min_{\x, \e} \|\e\|_1 \quad \subj \quad \b = A_k \x + \e.
\end{equation}
Where $A_k$ contains the training images for user $k$ as well as several vectors consisting of the
Jacobian of the test image $\b$ with respect to the transformation parameters, and $\x$ contains
both the current representation coefficients, as well as an update to the transformation parameters.

@@

	%Major steps:
	%while:
		%mu *= rho
		%lambda += mu(y - Ax - e)
		%while:
			%e <- shrink(y-Ax+lambda/mu, 1/mu)
			%x <- Apinvt'*(y-e+lamba/mu)
We summarize the entire ALM
algorithm as Algorithm~\ref{alg:alm}, where $\gamma$ denotes the
largest eigenvalue of the matrix $A^TA$. For the choice of parameter $\mu$, we take the same strategy as
in \cite{YangJ2009-pp} and set $\mu = 2m / \|\b\|_1$.
\begin{algorithm}[h]
\caption{\bf (Augmented Lagrange Multiplier Method for Global
Recognition)}
\begin{algorithmic}[1]
\STATE {\bf Input:} $\b \in \Re^m$, $A \in \Re^{m \times n}$,
$\x_1 = \mathbf{0}$, $\e_1 = \b$, $\blamda_1 =
\mathbf{0}$.
\WHILE{not converged ($k = 1,2,\ldots$)}
\STATE $\e_{k+1} = \textup{shrink}\left(\b - A\x_k + \frac{\blamda_k}{\mu}, \frac{1}{\mu}\right)$;
\STATE $t_1\leftarrow 1$, $\z_1 \leftarrow \x_k$, $\w_1 \leftarrow \x_k$;
\WHILE{not converged ($l = 1,2,\ldots$)}
\STATE $\w_{l+1} \leftarrow \textup{shrink}\left(\z_l +
\frac{1}{\gamma}A^T\left(\b - A\z_l - \e_{k+1} +
\frac{1}{\mu}\blamda_k\right), \frac{1}{\mu\gamma}\right)$;
\STATE $t_{l+1} \leftarrow \frac{1}{2}\left( 1 +
\sqrt{1+4t_l^2}\right)$;
\STATE $\z_{l+1} \leftarrow \w_{l+1} + \frac{t_l - 1}{t_{l+1}}(\w_{l+1} - \w_l)$;
\ENDWHILE
\STATE $\x_{k+1} \leftarrow \w_{l}$;
\STATE $\blamda_{k+1} \leftarrow \blamda_k + \mu (\b - A\x_{k+1} - \e_{k+1})$;
\ENDWHILE \STATE
{\bf Output:} $\x^* \leftarrow \x_k, \e^* \leftarrow \e_k$.
\end{algorithmic}
\label{alg:alm}
\end{algorithm}

@@

Optimization:
\begin{equation}
\min_{\x, \e} \|\e\|_1 \quad \subj \quad \b = A \x + \e.
\end{equation}

$A$ is $mxn$
$m = 5120$
$n = 32$

@@

\begin{algorithm}[thb]
\caption{\bf (Deformable Sparse Recovery and Classification for
Face Recognition)} \label{alg:deformable-src}
\begin{algorithmic}[1]
\STATE {\bf Input:} Frontal training images $A_i \in \Re^{m\times n_i}, i=1,2,\ldots,K$ for $K$ subjects,  a test image
$\b\in\Re^m$ and a deformation group $T$. 
\STATE
{\bf for} each subject $i$, 
\STATE \hspace{3mm} $\tau^{(0)}
\leftarrow I$. 
\STATE \hspace{3mm} {\bf while} not converged $(j=1,2,\ldots)$ {\bf do} 
\STATE \hspace{6mm}
$\tilde \b(\tau) \leftarrow \frac{\b \circ \tau}{\|\b \circ
\tau\|_2}$; \;\;\; $J \leftarrow  \frac{\partial}{\partial
\tau} \tilde\b(\tau)  \bigr|_{\tau^{(j)}} $;
%\STATE \hspace{6mm} $(\hat \x, \hat \e, \Delta \tau) \leftarrow \left\{\begin{array}{l} \arg \min_{\x,\e,\Delta \tau} \| \e \|_1 \\  \subj \; \b + J \Delta \tau = A_k \x + \e \end{array}\right.$
\STATE \hspace{6mm} $ \Delta \tau =  \arg\min \; \| \e \|_1  \;
\subj \; \tilde \b + J \Delta \tau = A_i \x + \e.$ 
\STATE
\hspace{6mm} $\tau^{(j+1)} \leftarrow \tau^{(j)} + \Delta
\tau$; 
\STATE \hspace{3mm} {\bf end while} \STATE {\bf end} \STATE Keep
the top $S$ candidates $k_1, \ldots, k_S$ with the smallest
residuals $\|\e\|_1$. \STATE Compute an average transformation
$\bar{\tau}$ from $\tau_{k_1}, \tau_{k_2}, \ldots, \tau_{k_S}$.
\STATE Update $\b \leftarrow \b \circ \bar{\tau}$ and $\tau_i
\leftarrow \tau_i \cdot \bar{\tau}^{-1}$ for $i = k_1, \dots,
k_S$. \STATE Set $A \leftarrow \big[ A_{k_1} \circ
\tau_{k_1}^{-1} \mid A_{k_2} \circ \tau_{k_2}^{-1} \mid \dots
\mid A_{k_S} \circ \tau_{k_S}^{-1} \big]$. \STATE Solve the
$\ell^1$-minimization problem:
$$\hat{\x} = \arg \min_{\x, \e} \| \x \|_1 + \|\e\|_1 \;\; \subj \;\; \b = A \x + \e.$$
\STATE Compute residuals $r_i(\b) = \| {\b} - {A}_i \, \delta_i(\hat{\x}) \|_2$ for $i = k_1, \dots, k_S$.
\STATE {\bf Output:} $\mbox{identity}(\b) = \arg\min_i r_i(\b)$.
\end{algorithmic}
\end{algorithm}

We summarize the entire ALM
algorithm as Algorithm~\ref{alg:alm}, where $\gamma$ denotes the
largest eigenvalue of the matrix $A^TA$. For the choice of parameter $\mu$, we take the same strategy as
in \cite{YangJ2009-pp} and set $\mu = 2m / \|\b\|_1$.
\begin{algorithm}[h]
\caption{\bf (Augmented Lagrange Multiplier Method for Global
Recognition)}
\begin{algorithmic}[1]
\STATE {\bf Input:} $\b \in \Re^m$, $A \in \Re^{m \times n}$,
$\x_1 = \mathbf{0}$, $\e_1 = \b$, $\blamda_1 =
\mathbf{0}$.
\WHILE{not converged ($k = 1,2,\ldots$)}
\STATE $\e_{k+1} = \textup{shrink}\left(\b - A\x_k +
\frac{1}{\mu}\blamda_k, \frac{1}{\mu}\right)$;
\STATE $t_1\leftarrow 1$, $\z_1 \leftarrow \x_k$, $\w_1 \leftarrow \x_k$;
\WHILE{not converged ($l = 1,2,\ldots$)}
\STATE $\w_{l+1} \leftarrow \textup{shrink}\left(\z_l +
\frac{1}{\gamma}A^T\left(\b - A\z_l - \e_{k+1} +
\frac{1}{\mu}\blamda_k\right), \frac{1}{\mu\gamma}\right)$;
\STATE $t_{l+1} \leftarrow \frac{1}{2}\left( 1 +
\sqrt{1+4t_l^2}\right)$;
\STATE $\z_{l+1} \leftarrow \w_{l+1} + \frac{t_l - 1}{t_{l+1}}(\w_{l+1} - \w_l)$;
\ENDWHILE
\STATE $\x_{k+1} \leftarrow \w_{l}$;
\STATE $\blamda_{k+1} \leftarrow \blamda_k + \mu (\b - A\x_{k+1} - \e_{k+1})$;
\ENDWHILE \STATE
{\bf Output:} $\x^* \leAftarrow \x_k, \e^* \leftarrow \e_k$.
\end{algorithmic}
\label{alg:alm}
\end{algorithm}

>>>>

%Motivate the paper by an overview about the scope and importance of $\ell_1$-minimization.
$\ell_1$-minimization is one of the most powerful tools to be added to the engineer's
toolbox in years.  It has strong theoretical motivation for both for use in recovering sparse
solutions to systems of linear equations, and has been shown to perform well even in the
presence of a combination of sparse and dense error.  Since solutions of linear systems
arise in practically every branch of engineering and there are very few application which
do {\em not} benefit from a robust error function, the applications are far to numerous to
itemize here.  
For many applications of $\ell_1$-minimization solving a system $b=Ax$, $A$ often has a
special structure that can be leveraged to compute portions of $A$ on the fly
as they are needed, speeding up perfomance of the minimization dramatically.
Unfortunately, there are many import applications where this does not hold true.
This paper targets applications where $\bb$ and $A$ contain
image data, and thus must be loaded every time they are used, as is the 
case for many problems in computer vision.
The techniques presented in this paper target these applications, as well
as others where the dictionary is dense and unstructured. $A$ must be loaded every time
it is used, and algorithm performance becomes highly dependent on memory bandwidth, rather
that arithmetic logic.

% Narrow focus for applications to Recognition Application
The primary motivating application for this paper is face recognition for access control applications.
Recently, advances in automatic face recognition have been made by re-casting
it as a sparse representation problem.  This core of this technique consists of
re-casting recognition as a sparse representation problem. This is done by  
stacking the test image into a vector $\bb$, stacking the training images for
all of the users into the columns of a matrix $A$, and solving the following
minimization problem:
\begin{equation}
\min_{\x, \e} \| \x \|_1 + \|\e\|_1 \quad \subj \quad \bb = A \x + \e.
\end{equation}
This optimization promotes a solution that is sparse both in $\x$ and $\e$. 
Sparsity in $\x$ arrises from the fact that only a subset of the training images (if any)
correspond to the user in the test image.  Sparsity in $\e$ arrises from the knowledge
that small occlusions in the test image will only corrupt a small subset of the pixels \cite{Wright2009-PAMI}.
The large coefficients in $\x$ will concentrate on the correct user, and can be used as the basis of a classifier.

For this idea to work, all of the images must be aligned. An iterative alignment
algorithm for this purpose was described as part of the \cite{Wagner2009-CVPR}, which was
the first paper to expand \cite{Wright2009-PAMI} into a functioning face recognition system.
This iterative alignment routine is based on repeatedly solving the following optimization problem:
\begin{equation}
\min_{\x, \e} \|\e\|_1 \quad \subj \quad \bb = A_k \x + \e.
\end{equation}
Where $A_k$ contains the training images for user $k$ as well as several vectors consisting of the
Jacobian of the test image $\bb$ with respect to the transformation parameters, and $\x$ contains
both the current representation coefficients, as well as an update to the transformation parameters.

@@

	%Major steps:
	%while:
		%mu *= rho
		%lambda += mu(y - Ax - e)
		%while:
			%e <- shrink(y-Ax+lambda/mu, 1/mu)
			%x <- Apinvt'*(y-e+lamba/mu)
We summarize the entire ALM
algorithm as Algorithm~\ref{alg:alm}, where $\gamma$ denotes the
largest eigenvalue of the matrix $A^TA$. For the choice of parameter $\mu$, we take the same strategy as
in \cite{YangJ2009-pp} and set $\mu_0 = 2m / \|\bb\|_1$. We set $\rho=1.5$.
\begin{algorithm}[h]
\caption{\bf (Augmented Lagrange Multiplier Method Used in Alignment Inner Loop)}
\begin{algorithmic}[1]
\STATE {\bf Input:} $\bb \in \Re^m$, $A \in \Re^{m \times n}$,
$\x_1 = \mathbf{0}$, $\blamda_1 = \mathbf{0}$.
\WHILE{not converged ($k = 1,2,\ldots$)}
\WHILE{not converged ($l = 1,2,\ldots$)}
\STATE $\e_{l+1} \leftarrow \textup{shrink}\left(\bb - A\x_l + \frac{\blamda_k}{\mu_k}, \frac{1}{\mu_k}\right)$;
\STATE $\x_{l+1} \leftarrow (A^\dagger)^T \left(\bb - \e_{l+1} + \frac{\blamda_k}{\mu_k} \right) $;
\ENDWHILE
\STATE $\blamda_{k+1} \leftarrow \blamda_k + \mu (\bb - A\x_{k+1} - \e_{k+1})$;
\STATE $\mu_{k+1} \leftarrow \rho\mu_k$;
\ENDWHILE \STATE
{\bf Output:} $\x^* \leftarrow \x_k, \e^* \leftarrow \e_k$.
\end{algorithmic}
\label{alg:alm}
\end{algorithm}

@@

Optimization:
\begin{equation}
\min_{\x, \e} \|\e\|_1 \quad \subj \quad \bb = A \x + \e.
\end{equation}

$A$ is $m \times n$
$m = 5120$
$n = 32$

@@

%\begin{algorithm}[thb]
%\caption{\bf (Deformable Sparse Recovery and Classification for
%Face Recognition)} \label{alg:deformable-src}
%\begin{algorithmic}[1]
%\STATE {\bf Input:} Frontal training images $A_i \in \Re^{m\times n_i}, i=1,2,\ldots,K$ for $K$ subjects,  a test image
%$\bb\in\Re^m$ and a deformation group $T$. 
%\STATE
%{\bf for} each subject $i$, 
%\STATE \hspace{3mm} $\tau^{(0)}
%\leftarrow I$. 
%\STATE \hspace{3mm} {\bf while} not converged $(j=1,2,\ldots)$ {\bf do} 
%\STATE \hspace{6mm}
%$\tilde \bb(\tau) \leftarrow \frac{\bb \circ \tau}{\|\bb \circ
%\tau\|_2}$; \;\;\; $J \leftarrow  \frac{\partial}{\partial
%\tau} \tilde\bb(\tau)  \bigr|_{\tau^{(j)}} $;
%%\STATE \hspace{6mm} $(\hat \x, \hat \e, \Delta \tau) \leftarrow \left\{\begin{array}{l} \arg \min_{\x,\e,\Delta \tau} \| \e \|_1 \\  \subj \; \bb + J \Delta \tau = A_k \x + \e \end{array}\right.$
%\STATE \hspace{6mm} $ \Delta \tau =  \arg\min \; \| \e \|_1  \;
%\subj \; \tilde \bb + J \Delta \tau = A_i \x + \e.$ 
%\STATE
%\hspace{6mm} $\tau^{(j+1)} \leftarrow \tau^{(j)} + \Delta
%\tau$; 
%\STATE \hspace{3mm} {\bf end while} \STATE {\bf end} \STATE Keep
%the top $S$ candidates $k_1, \ldots, k_S$ with the smallest
%residuals $\|\e\|_1$. \STATE Compute an average transformation
%$\bar{\tau}$ from $\tau_{k_1}, \tau_{k_2}, \ldots, \tau_{k_S}$.
%\STATE Update $\bb \leftarrow \bb \circ \bar{\tau}$ and $\tau_i
%\leftarrow \tau_i \cdot \bar{\tau}^{-1}$ for $i = k_1, \dots,
%k_S$. \STATE Set $A \leftarrow \big[ A_{k_1} \circ
%\tau_{k_1}^{-1} \mid A_{k_2} \circ \tau_{k_2}^{-1} \mid \dots
%\mid A_{k_S} \circ \tau_{k_S}^{-1} \big]$. \STATE Solve the
%$\ell^1$-minimization problem:
%$$\hat{\x} = \arg \min_{\x, \e} \| \x \|_1 + \|\e\|_1 \;\; \subj \;\; \bb = A \x + \e.$$
%\STATE Compute residuals $r_i(\bb) = \| {\bb} - {A}_i \, \delta_i(\hat{\x}) \|_2$ for $i = k_1, \dots, k_S$.
%\STATE {\bf Output:} $\mbox{identity}(\bb) = \arg\min_i r_i(\bb)$.
%\end{algorithmic}
%\end{algorithm}

%We summarize the entire ALM
%algorithm as Algorithm~\ref{alg:alm}, where $\gamma$ denotes the
%largest eigenvalue of the matrix $A^TA$. For the choice of parameter $\mu$, we take the same strategy as
%in \cite{YangJ2009-pp} and set $\mu = 2m / \|\bb\|_1$.
%\begin{algorithm}[h]
%\caption{\bf (Augmented Lagrange Multiplier Method for Global
%Recognition)}
%\begin{algorithmic}[1]
%\STATE {\bf Input:} $\bb \in \Re^m$, $A \in \Re^{m \times n}$,
%$\x_1 = \mathbf{0}$, $\e_1 = \bb$, $\blamda_1 =
%\mathbf{0}$.
%\WHILE{not converged ($k = 1,2,\ldots$)}
%\STATE $\e_{k+1} = \textup{shrink}\left(\bb - A\x_k +
%\frac{1}{\mu}\blamda_k, \frac{1}{\mu}\right)$;
%\STATE $t_1\leftarrow 1$, $\z_1 \leftarrow \x_k$, $\w_1 \leftarrow \x_k$;
%\WHILE{not converged ($l = 1,2,\ldots$)}
%\STATE $\w_{l+1} \leftarrow \textup{shrink}\left(\z_l +
%\frac{1}{\gamma}A^T\left(\bb - A\z_l - \e_{k+1} +
%\frac{1}{\mu}\blamda_k\right), \frac{1}{\mu\gamma}\right)$;
%\STATE $t_{l+1} \leftarrow \frac{1}{2}\left( 1 +
%\sqrt{1+4t_l^2}\right)$;
%\STATE $\z_{l+1} \leftarrow \w_{l+1} + \frac{t_l - 1}{t_{l+1}}(\w_{l+1} - \w_l)$;
%\ENDWHILE
%\STATE $\x_{k+1} \leftarrow \w_{l}$;
%\STATE $\blamda_{k+1} \leftarrow \blamda_k + \mu (\bb - A\x_{k+1} - \e_{k+1})$;
%\ENDWHILE \STATE
%{\bf Output:} $\x^* \leftarrow \x_k, \e^* \leftarrow \e_k$.
%\end{algorithmic}
%\label{alg:alm}
%\end{algorithm}
